---
title: "Storm Data Analysis"
author: "Seth Dobrin"
date: "January 15th, 2016"
output: html_document
---

Synopsis
---
This report analyses storm and natural disaster data from the United States leveraging NOAA data collected from 1950 thru 2011. Here we focus on the public health the economic impact caused by different types of events. This analysis finds that tornados have the most significant impact on both public health and economics. 

Data Processing
---

Load and preprocess the data
Store the URL for data
```{r Preprocess, echo = TRUE, tidy = TRUE}
fileURL <- c("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FstormData.csv.bz2")
docURL <- c("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf")
faqURL <- c("https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf")
```

Store filenames and paths as varriables
```{r File naming, echo = TRUE, tidy = TRUE}
fileName <- "~/GitHub/RepData/PA_2/repdata-data-stormData.csv"
zipFileName <- "~/GitHub/RepData/PA_2/repdata-data-stormData.csv.bz2"
docFileName <- "~/GitHub/RepData/PA_2/repdata-peer2_doc-pd01016005curr.pdf"
faqFileName <- "~/GitHub/RepData/PA_2/repdata-peer2_doc-NCDC Storm Events-FAQ Page.pdf"
```

Check to see if file exists. Download and extract only if it does not
```{r File management, echo = TRUE, tidy = TRUE}
library(R.utils, quietly = TRUE)
if (!file.exists(fileName)) {
        download.file(fileURL, destfile = zipFileName, method = "curl")
        bunzip2(file = zipFileName, fileName)
        date = Sys.Date()
}
```

Create data frame, reduce the complexity of the data by subsetting only the data needed for this report. Rename columns
```{r Preprocessing, echo = TRUE, tidy = TRUE, cache = TRUE}
library(dplyr) 
stormData <- read.csv(fileName, na.strings = " ")
stormData <- subset(stormData, select = c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG"))
stormData[is.na(stormData)] <- 0
colnames(stormData) <- c("Event", "Fatalities", "Injuries", "PropDmg", "CropDmg")
propDmg <- summarize(group_by(stormData, Event), sum(PropDmg))
colnames(propDmg) <- c("Event", "PropDmg")
cropDmg <- summarize(group_by(stormData, Event), sum(CropDmg))
colnames(cropDmg) <- c("Event", "CropDmg")
fatal <- summarize(group_by(stormData, Event), sum(Fatalities))
colnames(fatal) <- c("Event", "Fatalities")
injury <- summarize(group_by(stormData, Event), sum(Injuries)) 
colnames(injury) <- c("Event", "Injuries")
```

Order data frames and get summary assesments for reporting and further analysis. 
```{r Order data for reporting, echo = TRUE, cache = TRUE, tidy = TRUE}
summaryData <- Reduce(function(...) merge(..., all=TRUE), list(propDmg, cropDmg, fatal, injury))
top5PropDmg <- propDmg[with(propDmg, order(-PropDmg)),][1:5,] # order data frame and subset top 5
top5CropDmg <- cropDmg[with(cropDmg, order(-CropDmg)),][1:5,] # order data frame and subset top 5
top5Fatal <- fatal[with(fatal, order(-Fatalities)),][1:5,] # order data frame and subset top 5
top5Injury <- injury[with(injury, order(-Injuries)),][1:5,] # order data frame and subset top 5
economic <- Reduce(function(...) merge(..., all=TRUE), list(top5PropDmg, top5CropDmg))
economic <- mutate(economic, TotalDmg = (economic$PropDmg + economic$CropDmg))
health <- Reduce(function(...) merge(..., all=TRUE), list(top5Fatal, top5Injury))
health <- mutate(health, TotalHealth = (health$Fatalities + health$Injuries))
```

Results
---

Across the United States, which types of events are most harmful with respect to population health?


```{r Top public health events, echo = TRUE, cache = TRUE, tidy = TRUE}
health[with(health, order(-TotalHealth)),][1:5,]
```
Table 1: Summary assesments of the top public health impacts

```{r Plot top public health events, echo = TRUE, cache = TRUE, tidy = TRUE, fig.width = 8, fig.height = 5}
healthData <- melt(topHealth, id="Event")
healthData[is.na(healthData)] <- 0
ggplot(healthData, aes(x=Event, y=factor(value), fill = variable)) +
        geom_bar(stat="identity", position = "identity") + facet_wrap(~variable) + 
        theme(axis.text.x =
               element_text(size  = 10,
                            angle = 45,
                            hjust = 1,
                            vjust = 1))
```

Figure 1: Representation of the public health data showing the events contributing to injuries, fatalities and total (injuries + fatalities) health impact.

Across the United States, which types of events have the greatest economic consequences?

```{r Top economic events, echo = TRUE, cache = TRUE, tidy = TRUE}
economic[with(economic, order(-TotalDmg)),][1:5,]
```
Table 2: Summary assesments of economic impacts 


```{r Plot top economic events, echo = TRUE, cache = TRUE, tidy = TRUE, fig.width = 8, fig.height = 5}
library(data.table)
library(ggplot2)
damageData <- melt(topEconomic, id="Event")
damageData[is.na(damageData)] <- 0
ggplot(damageData, aes(x=Event, y=factor(value), fill = variable)) +
        geom_bar(stat="identity", position = "identity") + facet_wrap(~variable) + 
        theme(axis.text.x =
               element_text(size  = 10,
                            angle = 45,
                            hjust = 1,
                            vjust = 1))
```

Figure 2: Representation of the economic data showing the events contributing to property damage, crop damage and total (property damage + crop damage) economic impact.